Recent demography drives changes in linked selection across the maize genome

### 2015-10-30 manuscript.tex
incorporate
Incorporated Jeff's latest comments into the introduction.

%\jri{we need a starting paragraph on linked selection and demography. check out slotte 2014 briefings functional genomics for a great review of ideas. then we do a paragraph on why domesticates are great for this (modified version of your below)}
%\mbh{Yes, agree here...think the topics addressed in the manuscript need to be set up as generalizable and not just specific to domesticates.  Crops are a great system because we know a fair bit more about demography and the process of selection than in wild species, but they are not entirely unique.}

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to changes in genetic variabiligy \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. A classical approach for studying selection is to remove the genome-wide effects of demography and assume that selection will leave locus-specific evidence \cite{Wall2002, akey2009}. Conversely, demography is often researched by excluding potentially selected sites \cite{voight2005, luikart2003, gutenkunst2009}.

However, recent work has shown that demography and selection do not operate independently to determine genetic variability, in large part because linkage allows the effects of selection to be wide-ranging \cite{li2012,slotte2014}. For example, in \emph{Drosophila} the majority of the genome is
impacted by the effects of selection through linkage
\cite{sella2009}. One technique that has been utilized to estimate demographic parameters independently of selective parameters is to limit analyses to only sites far from genes\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is
difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the genome are not influenced by linked selection. Therefore, characterizing selection, demography or both, first requires an understanding of linked selection. It has already been established that patterns of linked selection are heavily influenced by population size ($N_e$) \cite{cutter2013, slotte2014}, but nearly all species including humans \cite{reich1998}, domesticates \cite{hyten2006, bovine2009}, ecological \cite{storfer2014}, and model species \cite{duchen2013} have experienced recent or ancient changes in $N_e$. Although much is known about how the long-term average $N_e$ affects linked selection, much less is understood about the immediate effects and dynamics of changing in $N_e$ on patterns of linked selection.

Because of its relatively simple demographic history and abundant availability of genomic tools, maize (\emph{Zea mays}) represents an excellent organism to study this phenomenon. Maize is a species of tremendous importance worldwide as both a model species \cite{strable2009} and a staple crop \cite{shiferaw2011}. Archaeological and genetic studies have established that maize domestication began in
Central Mexico at least 9,000 years bp
\cite{smith1995,matsuoka2002}, and involved a population bottleneck followed by recent expansion \cite{wright2005}. Because of this simple but dynamic demographic history, patterns of linked selection in wild and domesticated maize can be used to understand the effects of changing $N_e$ on linked selection. Teosinte, which remains extant throughout much of the Americas \cite{wilkes1967}, is the wild ancestor to maize. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified in maize. But despite all that is
known about maize domestication, the demographic parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication, and how linkage impacts the evidence and patterns of selection across the genome, remain unknown.

For this study, we begin by estimating the parameters of maize domestication based on whole-genome sequencing data. Next, we utilized these estimates to 1) investigate the relative importance of different forms of selection on genetic
variability in the recent and acient past; and 2) research the impact that demographic history, specifically changes in $N_e$, has had on linked selection in maize, and how this compares to the impact of a different demographic history in teosinte. Gaining an improved understanding of the impacts of a dynamic demography on whole-genome variability in maize will help to increase the understanding of these phenomena across all organisms, since changes in $N_e$ are pervasive.

>>>>

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to changes in genetic variability \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. A classical approach for studying selection is to remove the genome-wide effects of demography and assume that selection will leave locus-specific evidence \cite{Wall2002, akey2009}. Conversely, demography is often researched by excluding potentially selected sites \cite{voight2005, luikart2003, gutenkunst2009}.

However, demography and selection do not operate independently to determine genetic variability, in large part because linkage allows the effects of selection to be wide-ranging \cite{smith1974, li2012,slotte2014}. Linked selection can take the form of hitch-hiking, for which the frequency of neutral alleles changes as a result of positive selection at a physically linked site \cite{smith1974}, or background selection, where case diversity is reduced at loci physically linked to a site undergoing purifying selection \cite{charlesworth1993}. \emph{Drosophila} is an example of an organism for which linked selection is extremely important, as the majority of the genome has been impacted by it \cite{sella2009}. One technique that has been utilized to disentangle selection from demography is to limit analyses to only sites far from genes\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the genome are not influenced by linked selection. Therefore, characterizing selection, demography or both, first requires an understanding of linked selection.

One factor that heavily influences patterns of linked selection is population size ($N_e$) \cite{cutter2013, slotte2014}. but nearly all species including humans \cite{reich1998}, domesticates \cite{hyten2006, bovine2009}, ecological \cite{storfer2014}, and model species \cite{duchen2013} have experienced recent or ancient changes in $N_e$. Although much is known about how the long-term average $N_e$ affects linked selection, much less is understood about the immediate effects and dynamics of changing in $N_e$ on patterns of linked selection.

%\jri{main points are there. tempted to split into two paragraphs. 1) selection/drift not independent because of linked selection. very briefly explain it -- bgs vs purifying vs hard/soft. linked selection impacted by Ne. BGS models, hard sweep models, standing variaton (hermisson/pennings petrov/messer) etc. 2) but always assume long-term Ne. pops have gone through demography, this has potential to impact both selection and drift}

Because of its relatively simple demographic history and abundant availability of genomic tools, maize (\emph{Zea mays}) represents an excellent organism to study this phenomenon. Maize is a species of tremendous importance worldwide as both a model species \cite{strable2009} and a staple crop \cite{shiferaw2011}. Archaeological and genetic studies have established that maize domestication began in
Central Mexico at least 9,000 years bp
\cite{smith1995,matsuoka2002}, and involved a population bottleneck followed by recent expansion \cite{wright2005}. Because of this simple but dynamic demographic history, patterns of linked selection in wild and domesticated maize can be used to understand the effects of changing $N_e$ on linked selection. Teosinte, which remains extant throughout much of the Americas \cite{wilkes1967}, is the wild ancestor to maize. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified in maize. But despite all that is
known about maize domestication, the demographic parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication, and how linkage impacts the evidence and patterns of selection across the genome, remain unknown. 
For this study, we begin by estimating the parameters of maize domestication based on whole-genome sequencing data. Next, we utilized these estimates to 1) investigate the relative importance of different forms of selection on genetic
variability in the recent and acient past; and 2) research the impact that demographic history, specifically changes in $N_e$, has had on linked selection in maize, and how this compares to the impact of a different demographic history in teosinte. Gaining an improved understanding of the impacts of a dynamic demography on whole-genome variability in maize will help to increase the understanding of these phenomena across all organisms, since changes in $N_e$ are pervasive.
